Relational programming in Smalltalk

### 2019-03-07 booklet/article.tex
Apply fixes by Stef Ducasse.

\begin{abstract}
This booklet is an extended description of $\mu$Kanren, a relational
interpreter formerly implemented in Scheme and ported in Smalltalk for the sake
of understanding and, of course, fun. It enjoy simplicity and elegance despite
the complex nature of logic systems; moreover, using an uniform but powerful
language such as Smalltalk it is possible to appreciate and benefit from both
of them.
\end{abstract}

@@

It is a waste of time to repeat ourself on $\mu$Kanren's definition; on the
contrary, we believe interesting to look up and to watch over the many
implementation it has inspired since its seminal paper \citep{Hemann:muKanren}
and how programmers stretch the canonical implementation to take advantage of
their favorite programming language's features.

Certainly the book \citep{Friedman:Reasoned:Schemer} is the main reference
written in the same style of the other \emph{little books} of which Dan
Friedman coauthors them all; moreover, Byrd's dissertation \citep{Byrd:PhD}
shows advanced stuff and techniques while the repository
\citep{byrd:minikanren:repo} -- updated regularly by Byrd himself -- keeps
pointers to implementations, articles and presentations. All these references
uses the Scheme programming language, $\mu$Kanren is \emph{purely functional}
therefore any language that has the same power of the $\lambda$-calculus is a
candidate for implementing it, Scheme in particular; in parallel,
\citep{byrd:minikanren} has the same nature but uses Scheme macrology and in
some aspect it requires little effort to port to other languages that haven't
macros.

Another implementation using the \emph{Clojure} dialect of Lisp is
\citep{clojure:core.logic} which targets the JVM an provides constraints and
nominal logic programming; on the other hand, \citep{sullivan:microkanrenhs}
is a clean and concise implementation using Haskell, while many others simpler
versions can be found, written for the sake of understanding and, ultimately,
for fun.

Finally, your humble authors believe that $\mu$Kanren is a sophisticated
\textit{educational machine} because it concerns and mix together unification,
lazy streams, backtracking and optimization techniques; moreover, it allows
programmers to write their own version according to their personal taste and
style, ranging from implementations pretty imperative to ones built over
categories.  We repeat the exercise of writing it in \citep{Nocentini:kanrens}
targeting different languages, in particular
\begin{itemize}
\item using \emph{Python} we take advantage of native \verb|generator|s objects
and we provide a \emph{fair} enumeration strategy that the canonical one
hasn't; on the contrary, we experience computational limits because of
\texttt{RecursionError} exceptions. This implementation is complete and
parallels the one shown in the reference book, accompanied by a unit test suite
that covers \emph{all} questions explained in the book and some new
applications such as (i)~finding fixed points of a set of inference rules and
enumeration of classes of combinatorial objects.
\item using \emph{OCaml} the syntax of the language can be cleanly defined by
algebraic datatypes and the type system helps in the design of the required
operators.  While object-oriented languages make extensibility a relative
simple task, this implementation is harder unless we can make changes to basic
types, at least for what concerns the central \emph{unification} procedure.
All of this is preparatory to modify the tactics mechanism of the \emph{HOL
Light} theorem prover in order to introduce the relational paradigm in that
environment.
\end{itemize}
The current work describe a Smalltalk implementation of the basic system
with the aim to learn and experiment with a pure object environment, such as 
Pharo and Squeak. Starting small, the next example we show three different
kinds to perform an identity among \verb|SmallInteger| objects:

@@

Let us consider the object \Verb|false| whose responsibility is to represent
\textit{falsehood}, hence it should be a good candidate to provide us something
related to this from the \textit{relational} point of view,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
false asGoal
\end{minted}
namely, \Verb|false| can build for us an object that represents a relation that
cannot be satisfied at all responding to the message selector \Verb|asGoal| --
don't be scared by multiple names, we freely interchange the terms
\textit{goal} and \textit{relation} postponing to later sections the
description of their technical difference. Moreover, let us 
assume that the infrastructure that actually build results is given and implemented
by the \textit{oracle block} \texttt{B := [:g | g solutions all]}, we write our first test case:
\begin{minted}[fontsize=\footnotesize]{smalltalk}
GoalTest>>#testFail
    self 
        assert: false asGoal solutions all 
        equals: {}
\end{minted}
Now we get within the \Verb|False| class to add the implementation of the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
False>>#asGoal
    ^ Goal fail
\end{minted}
and of the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#fail
    ^ Fail new
\end{minted}
such that to say that \verb|Fail| denotes the impossibility to satisfy a relation
is the same to code the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Fail>>#onState: aState 
    ^ Chain bottom
\end{minted}
which will be used within the message \Verb|solutions| sent in the oracle block
\Verb|B| -- a lot remains to be explained, such as the \Verb|Chain| class and
streams in particular, all in what follows.

@@

We quickly review some mathematical backgrounds underlying logic systems;
in particular deductions are performed according to \emph{resolutions}

>>>>

\begin{abstract}
This booklet is an extended description of $\mu$Kanren, a relational
interpreter formerly implemented in Scheme and ported in Smalltalk for the sake
of understanding and, of course, fun. It enjoys simplicity and elegance despite
the complex nature of logic systems; moreover, using an uniform but powerful
language such as Smalltalk it is possible to appreciate and benefit from both
of them.
\end{abstract}

@@

It is a waste of time to repeat ourself on $\mu$Kanren's definition; on the
contrary, we believe interesting to look up and to watch over the many
implementations it has inspired since its seminal paper \citep{Hemann:muKanren}
and how programmers stretch the canonical implementation to take advantage of
their favorite programming language's features.

Certainly the book \citep{Friedman:Reasoned:Schemer} is the main reference
written in the same style of the other \emph{little books} of which Dan
Friedman coauthors them all; moreover, Byrd's dissertation \citep{Byrd:PhD}
shows advanced stuffs and techniques while the repository
\url{http://minikanren.org/} -- updated regularly by Byrd himself -- keeps
pointers to implementations, articles and presentations. All these references
use the Scheme programming language, $\mu$Kanren is \emph{purely functional}
therefore any language that has the same power of the $\lambda$-calculus is a
candidate for implementing it, Scheme in particular; in parallel, the
implementation with symbolic constraints has the same nature but uses Scheme
macrology and in some aspect it requires little effort to port to other
languages that haven't macros.

Another implementation using the \emph{Clojure} dialect of Lisp is
\citep{clojure:core.logic} which targets the JVM and provides constraints and
nominal logic programming; on the other hand, \citep{sullivan:microkanrenhs}
is a clean and concise implementation using Haskell, while many others simpler
versions can be found, written for the sake of understanding and, ultimately,
for fun.

Finally, your humble authors believe that $\mu$Kanren is a sophisticated
\textit{educational machine} because it concerns and mixes together unification,
lazy streams, backtracking and optimization techniques; moreover, it allows
programmers to write their own version according to their personal taste and
style, ranging from implementations pretty imperative to ones built over
categories.  We repeat the exercise of writing it in \citep{Nocentini:kanrens}
targeting different languages, in particular
\begin{itemize}
\item using \emph{Python} we take advantage of native \verb|generator|s objects
and we provide a \emph{fair} enumeration strategy that the canonical one
hasn't; on the contrary, we experience computational limits because of
\texttt{RecursionError} exceptions. This implementation is complete and
parallels the one shown in the reference book, accompanied by a unit test suite
that covers \emph{all} questions explained in the book and some new
applications such as (i)~finding fixed points of a set of inference rules and
enumeration of classes of combinatorial objects.
\item using \emph{OCaml} the syntax of the language can be cleanly defined by
algebraic datatypes and the type system helps in the design of the required
operators.  While object-oriented languages make extensibility a relative
simple task, this implementation is harder unless we can make changes to basic
types, at least for what concerns the central \emph{unification} procedure.
All of this is preparatory to modify the tactics mechanism of the \emph{HOL
Light} theorem prover in order to introduce the relational paradigm in that
environment.
\end{itemize}
The current work describes a Smalltalk implementation of the basic system
with the aim to learn and experiment with a pure object environment, such as 
Pharo and Squeak. Starting small, the next example we show three different
ways to express equality between \verb|SmallInteger| objects:

@@

Let us consider the object \Verb|false| whose responsibility is to represent
\textit{falsehood}, hence it should be a good candidate to provide us something
related to this from the \textit{relational} point of view,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
false asGoal
\end{minted}
namely, \Verb|false| can build for us an object that represents a relation that
cannot be satisfied at all responding to the message selector \Verb|asGoal| --
don't be scared by multiple names, we freely interchange the terms
\textit{goal} and \textit{relation} postponing to later sections the
description of their technical difference. Moreover, let us 
assume that the infrastructure that actually builds results is given and implemented
by the block \texttt{B := [:g | g solutions all]}, we write our first test case:
\begin{minted}[fontsize=\footnotesize]{smalltalk}
GoalTest>>#testFail
    self 
        assert: false asGoal solutions all 
        equals: {}
\end{minted}
Now we extend the \Verb|False| class with the following \Verb|asGoal| message
definition
\begin{minted}[fontsize=\footnotesize]{smalltalk}
False>>#asGoal
    ^ Goal fail
\end{minted}
and of the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#fail
    ^ Fail new
\end{minted}
such that to say that \verb|Fail| denotes the impossibility to satisfy a relation
is the same to code the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Fail>>#onState: aState 
    ^ Chain bottom
\end{minted}
which will be used within the message \Verb|solutions| sent in the block
\Verb|B| -- a lot remains to be explained, such as the \Verb|Chain| class and
streams in particular, all in what follows.

@@

We quickly review some mathematical backgrounds underlying logic systems;
in particular deductions are performed according to \emph{resolution} methods.
